---
title: "Comparing Peanut Production and Seed Size from Different Locations"
author: "Setiawan"
date: "September 21, 2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Importing Data

```{r data}
dta <- read.csv("peanut.csv")
dtasplit <- split(dta,dta$Location)
X_1 <- dtasplit$`1`
X_2 <- dtasplit$`2`

summary(dtasplit)
```

### Checking Normality

```{r norm}
## Checking Normality
qqnorm(X_1[, 3])
qqline(X_1[, 3])

qqnorm(X_1[, 5])
qqline(X_1[, 5])

qqnorm(X_2[, 3])
qqline(X_2[, 3])

qqnorm(X_2[, 5])
qqline(X_2[, 5])

X_bar_1 <- colMeans(X_1[, c(3, 5)])
X_bar_2 <- colMeans(X_2[, c(3, 5)])
S_1 <- var(X_1[, c(3, 5)])
S_2 <- var(X_2[, c(3, 5)])

n_1=n_2=6
p=2
```

Based on the Q-Q plots, we can safely assume that the data is normally distributed.


### Defining the Hypothesis
The hypothesis that will be tested is that the mean of peanut's total yields and seed sizes are the same for different locations.
By testing this hypothesis, we can determine whether mean of total yield and seed size varies by locations.

Testing H0: X_bar_1 - X_bar_2 = 0


### Statistical Testing
#### Assuming equal variances and Normal distribution

```{r eqvar}
S_po <- ((n_1 - 1) * S_1 + (n_2 - 1) * S_2) / (n_1 + n_2 - 2)

T2 <- (X_bar_1 - X_bar_2) %*% solve(S_po * (1 / n_1 + 1 / n_2)) %*% (X_bar_1 - X_bar_2)
p_value <- 1 - pf((n_1 + n_2 - p - 1) / ((n_1 + n_2 - 2) * p) * T2, p, n_1 + n_2 - p - 1)
qf(0.95, p, n_1 + n_2 - p - 1)
```
T2=1.84, p-value=0.468>0.05, therefore H0 is accepted.

#### Assuming unequal variances and Normal distribution

```{r uneqvar}
SS_1 <- S_1 %*% solve(S_1 / n_1 + S_2 / n_2) / n_1
SS_2 <- S_2 %*% solve(S_1 / n_1 + S_2 / n_2) / n_2
nu <- (p + p ^ 2) / ((sum(diag(SS_1 ^ 2)) + sum(diag(SS_1)) ^ 2) / n_1 + 
                       (sum(diag(SS_2 ^ 2)) + sum(diag(SS_2)) ^ 2) / n_2)
T2 <- (X_bar_1 - X_bar_2) %*% solve(S_1 / n_1 + S_2 / n_2) %*% (X_bar_1 - X_bar_2)
p_value <- 1 - pf((nu - p + 1) / (nu * p) * T2, p, nu - p + 1)
```
T2=1.84, p-value=0.457>0.05, therefore H0 is accepted.


### Conclusion
Both equal and unequal variance assumption end with the same result that H0 is accepted.
Therefore, mean of total yield and seed size does not vary with location.
