---
title: "Comparing Men and Women Oxygen Consumption"
author: "Setiawan"
date: "September 21, 2018"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
### Introduction
We have datasets of oxygen consumption from men and women while resting and while doing strenuous exercises. Using these datasets, we will try to see if there is a difference between men and women's oxygen consumptions. To do this, we are going to calculate the means of oxygen consumptions for mena and women during resting and doing strenuous exercises. After that, we are comparing the means of men and women for both cases.

### Importing Data

```{r data}
dta <- read.delim("oxygen.DAT", header = FALSE, sep = "")
colnames(dta) <- c("X_1", "X_2", "X_3", "X_4", "Gender")
attach(dta)
summary(dta)
```

### Calculating Means and Standard Deviations
```{r means}
X_bar_1 <- colMeans(dta[1:25, c(1, 3)])
X_bar_2 <- colMeans(dta[26:50, c(1, 3)])
S_1 <- var(dta[1:25, c(1, 3)])
S_2 <- var(dta[26:50, c(1, 3)])

n_1=n_2=25
p=2
```

### Defining the Hypothesis
The hypothesis that will be tested is that the mean of oxygen volumes while resting and strenous exercise between men and women are the same
Testing H0: X_bar_1 - X_bar_2 = 0


### Statistical Testing
#### Assuming Equal Variances
```{r eqvar}
S_po <- ((n_1 - 1) * S_1 + (n_2 - 1) * S_2) / (n_1 + n_2 - 2)

T2 <- (X_bar_1 - X_bar_2) %*% solve(S_po * (1 / n_1 + 1 / n_2)) %*% (X_bar_1 - X_bar_2)
p_value <- 1 - pf((n_1 + n_2 - p - 1) / ((n_1 + n_2 - 2) * p) * T2, p, n_1 + n_2 - p - 1)
cat('T2=',T2)
cat('p value=',p_value)
```
T2=81.7, p-value=7.17e-11<0.05, Therefore H0 is rejected.


#### Assuming Unequal Variances
```{r uneqvar}
SS_1 <- S_1 %*% solve(S_1 / n_1 + S_2 / n_2) / n_1
SS_2 <- S_2 %*% solve(S_1 / n_1 + S_2 / n_2) / n_2
nu <- (p + p ^ 2) / ((sum(diag(SS_1 ^ 2)) + sum(diag(SS_1)) ^ 2) / n_1 + 
                       (sum(diag(SS_2 ^ 2)) + sum(diag(SS_2)) ^ 2) / n_2)
T2 <- (X_bar_1 - X_bar_2) %*% solve(S_1 / n_1 + S_2 / n_2) %*% (X_bar_1 - X_bar_2)
p_value <- 1 - pf((nu - p + 1) / (nu * p) * T2, p, nu - p + 1)

X_bar_1[1] - X_bar_2[1] + c(-1, 1) * qt(1 - 0.05 / (2 * p), nu) * 
  sqrt(S_1[1, 1] / n_1 + S_2[1, 1] / n_2)
X_bar_1[2] - X_bar_2[2] + c(-1, 1) * qt(1 - 0.05 / (2 * p), nu) * 
  sqrt(S_1[2, 2] / n_1 + S_2[2, 2] / n_2)
cat('T2=',T2)

cat('p value=',p_value)
```
T2=81.7, p-value=9.69e-11<0.05, Therefore H0 is rejected.


#### Assuming Normal Distribution

```{r norm}
T2 <- (X_bar_1 - X_bar_2) %*% solve((S_1 * 1 / n_1) + (S_2 * 1 / n_2)) %*% (X_bar_1 - X_bar_2)
p_value <- 1 - pchisq((n_1 + n_2 - p - 1) / ((n_1 + n_2 - 2) * p) * T2, p)
qchisq(0.95, p)

cat('T2=',T2)

cat('p value=',p_value)
```
T2=81.7, p-value=2.06e-09<0.05, Therefore H0 is rejected.



### Conclusion
The 3 assumptions end with same result, which is H0 is rejected.
The H0 that the mean of oxygen volumes while resting and strenous exercise between men and women are the same is rejected, and therefore there is no evidence that the mean of oxygen consumption between men and women are the same.
